---
title: "Data generation"
subtitle: "A simulated dataset for package functionality demonstration"
author: "Janick Weberpals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 100,
  fig.width = 6,
  fig.height = 4.5
  )
```

```{r setup}
library(smdi)
library(dplyr)

# some global simulation parameters
seed_value <- 42
n <- 1000
```

# `smdi` dataset background

To get acquainted with the functionality and usage of the `smdi` package, a simulated toy dataset is provided. The dataset is an exemplary low-dimensional electronic health records (EHR) dataset depicting a cohort of `r formatC(n, format = "fg", big.mark = ",")` lung cancer patients.

## Exposure and outcome

We assume that we are interested in studying the comparative effectiveness of two antineoplastic systemic drug treatment regimens (`exposure (0/1)`) on the time to death due to any reason (`overall survival`). The anticipated time of follow-up is truncated to 2 years.

The desired strength of effectiveness of the **exposure of interest** is defined with a hazard ratio of 0.7, i.e. the hazard of death due to any reason is decreased by 30% per unit of time under the proportional hazards assumption which is fulfilled for this dataset.

## Confounders

We further assume that there is a biomarker (`EGFR`) with rare to moderate prevalence (\~10-30%) which serves as a strong confounding covariate. Therefore, EGFR is strongly associated with the probablity of treatment initiation (OR \~ 0.7) and the outcome (HR \~ 0.5), indicating that patients with an alteration in the EGFR gene are more likely to get treated with the exposure of interest and benefit signficiantly more than wild-type EGFR patients.

Despite the low dimensionality, the dataset is simulated as realistically as possible with varying strengths of associations between covariates and treatment initiation and the outcome.

## Overview covariates/confounder structure

To get an overview, this table provides a summary of the different covariate-exposure-outcome correlations.

| Variable                  | Description                                      | Associated with exposure | Associated with outcome |
|------------------|------------------|------------------|------------------|
| age_num                   | Age at baseline (continuous)                     | Yes                      | Yes                     |
| female_cat                | Female gender (binary)                           | Yes                      | Yes                     |
| ecog_cat                  | ECOG performance score (binary 0 vs 1)           | Yes                      | Yes                     |
| smoking_cat               | Smoker vs. non-smoker at baseline (binary)       | Yes                      | Yes                     |
| egfr_cat                  | EGFR alteration (binary)                         | Yes                      | Yes                     |
| kras_cat                  | Alteration in KRAS gene (binary)                 | Yes                      | Yes                     |
| histology_is_squamous_cat | Tumor histology squamous vs nonsquamous (binary) | No                       | Yes                     |
| ldl_num                   | LDL cholesterol level (continuous)               | No                       | Yes                     |
| rr_num                    | Average blood pressure in mmHg (continuous)      | No                       | Yes                     |
| bpm_num                   | Average heart rate in bpm (continuous)           | No                       | Yes                     |

: Overview of simulated lung cancer dataset

# Simulation of covariates and exposure

```{r, covar_generation}
set.seed(seed_value)

# start with basic dataframe, covariates and their association with exposure
sim_covar <- tibble::tibble(
  exposure = rbinom(n = n, size = 1, prob = 0.4),
  age_num = rnorm(n, mean = 64 - 5.5*exposure, sd = 13.7),
  female_cat = rbinom(n, size = 1, prob = 0.39 - 0.12*exposure),
  ecog_cat = rbinom(n, size = 1, prob = 0.63 - 0.08*exposure),
  smoking_cat = rbinom(n, size = 1, prob = 0.15 + 0.1*exposure), 
  egfr_cat = rbinom(n, size = 1, prob = 0.10 + 0.2*exposure), 
  kras_cat = rbinom(n, size = 1, prob = 0.07 + 0.05*exposure),
  histology_is_squamous_cat = rbinom(n, size = 1, prob = 0.2),
  ldl_num = rnorm(n, mean = 115.1, sd = 41.1), # cholesterol
  rr_num = rnorm(n, mean = 123.7, sd = 20.6), # systolic blood pressure
  bpm_num = rnorm(n, mean = 74.2, sd = 11.7) # heart rate
  ) %>%  
  # bring data in right format
  dplyr::mutate(across(ends_with("num"), as.numeric)) %>% 
  dplyr::mutate(across(ends_with("num"), function(x) round(x, digits = 2)))
```

In the first step, we create a dataset with `r formatC(nrow(sim_covar), format = "fg", big.mark = ",")` patients and `r ncol(sim_covar)` variables with the following distributions.

```{r, distributions_covars}
sim_covar %>% 
  gtsummary::tbl_summary(by = "exposure") %>% 
  gtsummary::add_difference()
```

The following table illustrates the odds of exposure assignment.

```{r, pr(exposure)_tbl}
exposure_form <- as.formula(paste("exposure ~ ", paste(colnames(sim_covar %>% dplyr::select(-exposure)), collapse = " + ")))

exposure_fit <- stats::glm(
  exposure_form,
  data = sim_covar,
  family = "binomial"
  )

exposure_fit %>% 
  gtsummary::tbl_regression(exponentiate = T)
```

Fitting a generalized linear model and assessing the probablity of treatment assignment, the above constellation of odds results in the following simulated distributions depicting propensities of treatment initiation (aka propensity scores).

```{r, pr_treatment_assignment, fig.cap="Treatment assignment probabilities."}
# compute propensity score
exposure_plot <- sim_covar %>% 
  dplyr::mutate(ps = fitted(exposure_fit))

# plot density
exposure_plot %>% 
  ggplot2::ggplot(ggplot2::aes(x = ps, fill = factor(exposure))) +
  ggplot2::geom_density(alpha = .5) +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = "Pr(exposure)",
    y = "Density",
    fill = "Exposed"
  )
```

## Simulate time-to-event

Next, we simulate a time-to-event outcome for `overall survival`. For this, the `simsurv` package is used with the following assumptions:

-   Parametric event times follow an exponential distribution
-   Max event times: 2 years of follow-up
-   Event times depend on some (time-fixed/baseline) covariate effects as listed in the table above
-   beta coefficients for the outcome model as defined in the following
    -   HR(exposure) = 0.7
    -   HR(EGFR) = 0.5 (strong prognostic factor)

```{r, betas_outcome_generation}
betas_os <- c(
  exposure = log(0.7),
  age_num = log(1.05),
  female_cat = log(0.9),
  ecog_cat = log(1.2),
  smoking_cat = log(1.3),
  egfr_cat = log(0.5),
  kras_cat = log(0.95),
  histology_is_squamous_cat = log(1.15),
  ldl_num = log(1.01),
  rr_num = log(1.05),
  bpm_num = log(1.005)
  )

betas_os %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "Covariate") %>% 
  gt::gt()
```

```{r, outcome_generation, message=FALSE}
set.seed(seed_value)

sim_df <- sim_covar %>% dplyr::bind_cols(
  simsurv::simsurv(
    dist = "exponential",
    lambdas = 0.00005,
    betas = betas_os,
    x = sim_covar,
    maxt = 2
    )
  ) %>% 
  dplyr::mutate(id = paste0("ENROL", id)) %>% 
  dplyr::relocate(id, .before = exposure)
```

### Kaplan-Meier estimates

The simulation resulted in the following crude Kaplan-Meier estimates for 1-yr and 2-ry survival. Estimates belown are shown only for overall and by exposure and EGFR biomarker for overall survival.

```{r, km_estimates}
km_overall <- survival::survfit(survival::Surv(eventtime, status) ~ 1, data = sim_df)
km_exposure <- survival::survfit(survival::Surv(eventtime, status) ~ exposure, data = sim_df)
km_egfr <- survival::survfit(survival::Surv(eventtime, status) ~ egfr_cat, data = sim_df)

gtsummary::tbl_survfit(
  list(km_overall, km_exposure, km_egfr),
  times = c(1, 2),
  label_header = "**{time} Year**"
  )
```

### Kaplan-Meier curves

#### By exposure

```{r, km_plot_exposure}
km_exposure <- survival::survfit(survival::Surv(eventtime, status) ~ exposure, data = sim_df)

survminer::ggsurvplot(
  km_exposure, 
  data = sim_df,
  conf.int = TRUE,
  surv.median.line = "hv",
  palette = "jco",
  legend.labs = c("Comparator", "Exposure of interest")
  )
```

#### By EGFR

```{r, km_plot_egfr}
km_egfr <- survival::survfit(survival::Surv(eventtime, status) ~ egfr_cat, data = sim_df)

survminer::ggsurvplot(
  km_egfr, 
  data = sim_df,
  conf.int = TRUE,
  surv.median.line = "hv",
  palette = "jco",
  legend.labs = c("EGFR wild-type", "EGFR alteration")
  )
```

### Cox proportional hazards

Finally, the simulated dataset results in the following hazard ratios (HR)

```{r, Cox_estimates}
cox_lhs <- "survival::Surv(eventtime, status)"
cox_rhs <- paste(colnames(sim_covar), collapse = " + ")
cox_form = as.formula(paste(cox_lhs, "~ exposure +", cox_rhs))
  
cox_fit <- survival::coxph(cox_form, data = sim_df)

cox_fit %>% 
  gtsummary::tbl_regression(exponentiate = T)
```

# Overview final simulated dataset

Let's have a look at the final dataset.

```{r, distributions_covars_final}
sim_df %>% 
  dplyr::select(-id) %>% 
  gtsummary::tbl_summary(by = "exposure") %>% 
  gtsummary::add_difference()
```

# Export data

Exporting the data to `data/`, so it can be used for educative purposes and for playing around with the package.

```{r, export, message=FALSE}
smdi_data <- sim_df
usethis::use_data(smdi_data, overwrite = TRUE)
```
