---
title: "Multivariate missingness"
author: "Janick Weberpals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate missingness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.width = 6,
  fig.height = 4.5
  )
```

```{r setup}
library(smdi)
library(gt)
suppressPackageStartupMessages(library(dplyr))
```

# Multivariate missing data in `smdi`

In this article, we want to briefly highlight two aspect regarding **multivariate missingness**: 

1. How does `smdi` handle multivariate missingness 

2. What is the link between missing data patterns and missing data mechanisms and how this may affect the behavior and performance of the `smdi` functionality.

## Established taxonomies

In general, there are two classic established missing data taxonomies:

* Mechanisms: Missing completely at random (MCAR), at random (MAR) and not at random (MNAR)

* Patterns: Monotone versus Non-monotone missingness
