---
title: "Routine structural missing data diagnostics"
author: "Janick Weberpals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Routine structural missing data diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 100,
  fig.width = 6,
  fig.height = 4.5
  )
```

```{r setup}
library(smdi)
suppressPackageStartupMessages(library(dplyr))
```

# `smdi` main functionality

## Illustrative dataset

To illustrate the usage of the `smdi` package main functions, we use the `smdi_data` dataset which is an example dataset that comes bundled with the package. In brief, the `smdi_data` dataset consists of a simulated lung cancer cohort with a fictional comparison of two antineoplastic systemic therapy regimens and a time-to-event outcome. More information on the underlying dataset is given in the [Data generation](https://janickweberpals.gitlab-pages.partners.org/smdi/articles/data_generation.html) article.

```{r}
smdi::smdi_data %>% 
  dplyr::glimpse()
```

The dataset consists of `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` patients and `r ncol(smdi_data)` variables with `exposure` representing the two treatment regimens under comparison and `status` and `eventtime` the vital status and censoring time, respectively.



