---
title: "Get started with smdi"
author: "Janick Weberpals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with smdi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.width = 6,
  fig.height = 4.5
  )
```

```{r setup}
library(smdi)
```

# `smdi` - the flagship function

The `smdi` flagship function is `smdi_diagnose()` which calls multiple sub-functions which are also accessible separately. `smdi_diagnose()` builds on theoretical concepts developed and validated in a comprehensive simulation study as part of the FDA Sentinel Innovation Center workstream [Approaches to Handling Partially Observed Confounder Data From Electronic Health Records (EHR) In Non-randomized Studies of Medication Outcomes](https://www.sentinelinitiative.org/methods-data-tools/methods/approaches-handling-partially-observed-confounder-data-electronic-health).

```{r}
smdi_diagnose(
  data = smdi_data,
  covar = NULL,
  model = "cox",
  form_lhs = "Surv(eventtime, status)"
  )
```
