---
title: "Get started with smdi"
author: "Janick Weberpals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with smdi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.width = 6,
  fig.height = 4.5
  )
```

```{r setup}
library(smdi)
library(gt)
```

# `smdi_diagnose()` - the flagship function

The `smdi` main function is `smdi_diagnose()` which calls all three group diagnostics, all of which are also accessible individually. 

`smdi_diagnose()` builds on theoretical concepts developed and validated in a comprehensive simulation study based on the [workstream:]((https://www.sentinelinitiative.org/methods-data-tools/methods/approaches-handling-partially-observed-confounder-data-electronic-health)) 

*Approaches to Handling Partially Observed Confounder Data From Electronic Health Records (EHR) In Non-randomized Studies of Medication Outcomes*.

A most minimal example could look like this (if you want to accept all of the default parameters).

```{r}
diagnostics <- smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)"
  )

gt(diagnostics$smdi_tbl) %>% 
  tab_footnote(
    footnote = "Group 1 diagnostic: Differences in patient characteristics between patients with and without covariate",
    locations = cells_column_labels(
      columns = c(asmd_median, hotteling_p)
      )
    ) %>% 
  tab_footnote(
    footnote = "Group 2 diagnostic: Ability to predict missingness",
    locations = cells_column_labels(
      columns = rf_auc
      )
    ) %>% 
  tab_footnote(
    footnote = "Group 3 diagnostic: Assessment if missingness is associated with the outcome (crude, adjusted)",
    locations = cells_column_labels(
      columns = c(estimate_crude, estimate_adjusted)
      )
    )
```

```{r}
diagnostics$p_little
```

